---
title: "Day 2 - Hands On 2"
author: "Alejandro Martinez Hernandez"
date: "2026-07-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Tutorial for running a linear model**
Running a linear model in R involves a few simple steps:

### 1. Importing the data
In this case, we will use Dataset 3, which is stored in a CSV file.
```{r}
data <- read.table("DATA_SET_REFERENCE_3.csv", 
                   header = TRUE,
                   row.names = 1,
                   sep=','
)
```

### 2. Cleaning the data
We need to remove illogical values (negative values, in this case) as well as outliers. The threshold for an outlier is defined as follows:

$$\text{Outlier} = \text{median} \pm 3 \times \text{sd}$$
```{r}
## Load the function
source("my_functions.R")

## Use the cleaner
data_f <- data_cleaner(data)
head(data_f)
```


### 3.Creating the linear model
We use the lm() function to define the relationship we want to analyze. For this exercise, we will try to explain Exercise levels based on LDL levels.
```{r}
linear_model <- lm(Exercise ~ LDL_levels, data_f)
summary(linear_model)
plot(linear_model)
```


### 4. Checking residual errors
To validate our model, in addition to checking the p-values of the intercept $\beta_0$ and slope $\beta_1$ (both <2e-16 ), we must verify that the residuals follow a normal distribution. We can easily check this using the Shapiro-Wilk test on the model's residuals.
```{r}
test_n <- shapiro.test(residuals(linear_model))
print(test_n)
```


### 5. Conclusion
In this scenario, the p-values from the lm() output for $\beta_0 = 161.21222$ and $\beta_1 = 0.90699$ are extremely small (<2e-16), meaning both parameters are highly significant. Furthermore, the Shapiro-Wilk p-value for the residuals is relatively large (0.1571), so we cannot reject the assumption of normality. Based on this evidence, our model adequately explains the data.


 