---
title: "FINAL EVALUATION"
subtitle: "PART 1: Analysis of a small data-set | BESE-300"
bibliography: references.bib
citeproc: true
author: 
  - name: "Alejandro Martinez Hernandez | KAUST ID: 219559"
    affiliations:
      - name: "King Abdullah University of Science and Technology
      "
number-sections: true
format: 
  typst:
    columns: 2
    fontsize: 10pt
    margin:
      top: 15mm
      bottom: 15mm
      left: 15mm
      right: 15mm
knitr:
  opts_chunk:
    fig.align: center
editor: visual
---

# Data-set used in the analysis

The present work analyses the fictional data-set 'DATA_SET_REFERENCE_EVAL.csv', which presents 400 measurements of HDL and LDL in patients across two planets (Earth and Venus) from two populations (tall and small), as seen in @tbl-summary-py. No further information was given about the dataset due to it being fictional; nonetheless, in a real one, the source must be trustworthy with traceable methods.

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| include: false
library(here)
# LOAD THE DATA FRAME
#df_1 <- read.csv("../data/Part_1/DATA_SET_REFERENCE_EVAL.csv", row.names = 1)
df_1 <- read.csv(here("data", "Part_1", "DATA_SET_REFERENCE_EVAL.csv"), row.names = 1)
```

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-summary-py
#| tbl-cap: "Fictional Data-Set. First 6 entries"
#| echo: false
#head(df_1) # Check data
knitr::kable(head(df_1, n = 3L))
```

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| echo: false
#| include: false
print(paste("Number of rows:", nrow(df_1))) # Check number of rows
print(paste("Number of columns:", ncol(df_1))) # Check number of columns
print(paste("Column names:", paste(colnames(df_1), collapse = ", "))) 
print(paste("Unique values inside 'Size' column:", 
            paste(unique(df_1$Size), collapse = ", ")))
print(paste("Unique values inside 'Planet' column:", 
            paste(unique(df_1$Planet), collapse = ", ")))
print(paste("Number of 'Venus' patients:", sum(df_1$Planet == "Venus", na.rm = TRUE)))
print(paste("Number of 'Earth' patients:", sum(df_1$Planet == "Earth", na.rm = TRUE))) 
print(paste("Number of tall patients:", sum(df_1$Size == "Tall", na.rm = TRUE))) 
print(paste("Number of small patients:", sum(df_1$Size == "Small", na.rm = TRUE)))
```

## Overview

- **Total Records***:* 400 patients.

- **Locations:** Distributed nearly evenly across two planets (204 on Venus, 196 on Earth).

- **Demographics:** Exactly split by size (200 Tall, 200 Small).

- **Columns:**

  - **Patient:** Unique identifier (Patient1 to Patient400).

  - **HDL (High-Density Lipoprotein levels):** Numerical variable (mg/dL).

  - **LDL (Low-Density Lipoprotein levels) :** Numerical variable (mg/dL).

  - **Size:** Categorical variable (Tall or Small).

  - **Planet:** Categorical variable (Venus or Earth).

Both columns, HDL and LDL, contain negative values (something impossible) and one NA value each, telling us that the data must be cleaned. Apart from that, all the other columns seem to be in order, without typos or strange categories; therefore, most of the attention will be centered on the columns HDL and LDL.

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-summary2-py
#| tbl-cap: "Data-Set Summary"
#| echo: false
#| include: false
# summary(df_1)
knitr::kable(summary(df_1))
```

# Curation of the data

## Negative and NA values

It was decided to remove each patient who has a negative or NA value. For this particular data-set, both negative and NA values for HDL and LDL corresponded to only 2 patients as seen in @tbl-negative-NA-py.

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-negative-NA-py
#| tbl-cap: "Negative and NA Values"
#| echo: false
# Checking the patient to eliminate
n_idx_hdl <- df_1$HDL <= 0 | is.na(df_1$HDL)
n_idx_ldl <- df_1$LDL <= 0 | is.na(df_1$LDL)

df_f_o <- df_1[which(n_idx_hdl | n_idx_ldl), ]
#head(df_f_o)
knitr::kable(head(df_f_o))
```

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| include: false
## Removing negative values
df_2 <- data.frame(df_1) # Security copy

### Idx 1: Delete individuals with HDL < 0
idx_hdl <- df_2$HDL >= 0 & !is.na(df_2$HDL)

### Idx 2: Delete individuals with LDL < 0
idx_ldl <- df_2$LDL >= 0 & !is.na(df_2$LDL)

### Cleaning with filters
df_3 <- df_2[idx_hdl & idx_ldl, ] # Here we remove 2 values.
print(paste("Number of patients after removing negative and NA values: ", nrow(df_3)))
```

## Outliers

Outliers were defined using a threshold of three standard deviations from the mean. For each numerical column in the dataset, we calculated the upper limit (@eq-outlier_1) and lower limit (@eq-outlier_2) as follows:

$$\text{Upper Limit} = \text{mean} + 3 \times \text{SD}$$ {#eq-outlier_1}

$$\text{Lower Limit} = \text{mean} - 3 \times \text{SD}  $$ {#eq-outlier_2}

However, after applying these filters, no values were excluded, resulting in the same 398 observations remaining from the 'Negative and NA values' step.

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| include: false
## Identifying and removing outliers
## Define outlier: > mean - 3*sd or > mean + 3*sd
max_limit_LDL <- mean(df_3$LDL, na.rm = T) + 3*sd(df_3$LDL, na.rm = T)
min_limit_LDL <- mean(df_3$LDL, na.rm = T) - 3*sd(df_3$LDL, na.rm = T)
max_limit_HDL <- mean(df_3$HDL, na.rm = T) + 3*sd(df_3$HDL, na.rm = T)
min_limit_HDL <- mean(df_3$HDL, na.rm = T) - 3*sd(df_3$HDL, na.rm = T)

## Creating filters
clean_LDL <- min_limit_LDL < df_3$LDL & 
  max_limit_LDL > df_3$LDL # Only TRUE

clean_HDL <- min_limit_HDL < df_3$HDL & 
  max_limit_HDL > df_3$HDL 

## Cleaning with filters
df_4 <- df_3[clean_LDL & clean_HDL, ]
sum(df_4$HDL == df_3$HDL)
```

# Data exploration

## Normalization and balance

Although measurement units were not provided, HDL and LDL are assumed to be in milligrams per deciliter (mg/dL) based on the magnitude of the numbers and a comparison with average population values reported by @tsao2023heart (\~111 mg/dL and \~53 mg/dL, respectively). All values in the dataset are biologically possible, though some are exceptionally rare. While literature ranges are often used to filter extreme values or check data quality, this strategy was not applied here given the fictional nature of the dataset.

The HDL and LDL distributions were explored using histograms and Q-Q plots (@fig-qqplot-ldl & @fig-qqplot-hdl). The visualizations suggest that both variables follow a normal distribution (later confirmed by a statistical test); therefore, there was no need for data transformation. Finally, a cross-tabulation (@tbl-balance) confirmed that the comparison groups are balanced.

## Comparing Categorical and Continuos Data

The numerical variables (HDL and LDL) were grouped by Size and Planet and visualized using boxplots. There appeared to be no significant difference in LDL or HDL when grouping by planet; however, analyzing @fig-ldl-size and @fig-hdl-size suggests a notable trend when grouping by size, indicating a potential difference in HDL and LDL levels between tall and small individuals.

```{r}
#| include: false
library(knitr)
library(dplyr)

balance_table <- df_4 %>%
  group_by(Planet, Size) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(Percent = round(100 * n / sum(n), 1))
```

```{r}
#| label: tbl-balance
#| fig-cap: "Sample sizes by Planet and Size group"
#| echo: false

kable(balance_table, align = "c")
```

```{r}
#| label: fig-hts-1
#| fig-cap: "Histogram of LDL levels"
#| include: false
#| fig-width: 5
#| fig-height: 5

hist(df_4$LDL,
    main="")
```

```{r}
#| label: fig-hts-2
#| fig-cap: "Histogram of HDL levels"
#| include: false
#| fig-width: 5
#| fig-height: 5

hist(df_4$HDL,
    main="")
```

```{r}
#| label: fig-qqplot-ldl
#| fig-cap: "Q-Q Plot of LDL levels"
#| echo: false
#| fig-width: 4
#| fig-height: 4

qqnorm(df_4$LDL, pch = 1, frame = FALSE, main="")
qqline(df_4$LDL, col = "steelblue", lwd = 2)
```

```{r}
#| label: fig-qqplot-hdl
#| fig-cap: "Q-Q Plot of HDL levels"
#| echo: false
#| fig-width: 4
#| fig-height: 4

qqnorm(df_4$HDL, pch = 1, frame = FALSE, main="")
qqline(df_4$HDL, col = "steelblue", lwd = 2)
```

## Comparing Continuos Data

To explore the correlation between the continuous variables, LDL and HDL, a scatterplot was created (@fig-scatterplot_1), using color to indicate Size (Tall or Small) and shape for Planet (Earth and Venus). The plot reveals a negative trend: regardless of planet, small patients tend to have lower LDL and higher HDL levels, while the inverse is true for tall individuals.

```{r}
#| label: fig-ldl-size
#| fig-cap: "LDL levels by Size group"
#| echo: false
#| fig-width: 5
#| fig-height: 5

# Colorblind-friendly palette (Okabe-Ito), one color per Size group
group_colors <- c("Small" = "#E69F00", "Tall" = "#56B4E9")

boxplot(LDL ~ Size, 
        data = df_4, 
        col = group_colors,
        density = c(25, 70),   # different shading density = second visual cue, not just color
        pch = 16,
        ylab = "LDL")

```

```{r}
#| label: fig-hdl-size
#| fig-cap: "B. HDL levels by Size group"
#| echo: false
#| fig-width: 5
#| fig-height: 5
boxplot(HDL ~ Size, 
        data = df_4, 
        col = group_colors,
        density = c(25, 70), 
        pch = 16,
        ylab = "HDL")
```

```{r}
#| label: fig-ldl-planet
#| fig-cap: "LDL levels by Planet group"
#| include: false
#| fig-width: 5
#| fig-height: 5

# Colorblind-friendly palette (Okabe-Ito), one color per Size group
group_colors <- c("Small" = "#E69F00", "Tall" = "#56B4E9")

boxplot(LDL ~ Size, 
        data = df_4, 
        col = group_colors,
        density = c(25, 70),   # different shading density = second visual cue, not just color
        pch = 16,
        ylab = "LDL")
```

```{r}
#| label: fig-hdl-planet
#| fig-cap: "HDL levels by Planet group"
#| include: false
#| fig-width: 5
#| fig-height: 5
boxplot(HDL ~ Size, 
        data = df_4, 
        col = group_colors,
        density = c(25, 70), 
        pch = 16,
        ylab = "HDL")
```

```{r}
#| echo: false
#| label: fig-scatterplot_1
#| fig-cap: "HDL vs LDL by Size (color) and Planet (shape)"
#| fig-width: 5
#| fig-height: 6
# Comparing two continuos values
df_4$Group <- interaction(df_4$Planet, df_4$Size, sep = "-") # Combined group factor
# Levels will be: Earth-Small, Venus-Small, Earth-Tall, Venus-Tall (alphabetical/factor order)
group_levels <- levels(df_4$Group)
cb_colors <- c("#E69F00", "#0072B2")  # for Size: Small, Tall
shapes <- c(16, 17)                    # for Planet: Earth = circle, Venus = triangle

plot(df_4$LDL, df_4$HDL,
     col = cb_colors[as.factor(df_4$Size)],
     pch = shapes[as.factor(df_4$Planet)],
     xlab = "LDL", ylab = "HDL")

legend("topright",
       legend = c(levels(as.factor(df_4$Size)), levels(as.factor(df_4$Planet))),
       col = c(cb_colors, "black", "black"),
       pch = c(19, 19, shapes),
       pt.cex = 1, bty = "n")
```

```{r}
#| echo: false
write.csv(df_4, "../data/Part_1/CLEAN_DATA_SET_REFERENCE_EVAL.csv", row.names = FALSE)
```

# Statistics used to validate associations

## Normality Analysis

Shapiro-Wilk was used to analyze normality in this set due to its size (398 samples \< 5000) [@royston1982extension].

## 1 Analysis of 2 Continuous Variables

@fig-scatterplot_1 suggested a potential negative correlation between HDL and LDL. To formally evaluate this relationship, the normality of both variables was first assessed using the Shapiro-Wilk test (@tbl-normality). The results for both variables yielded $p-values \> 0.05$, failing to reject the null hypothesis. Because the assumption of normality was met, Pearson's parametric correlation test was selected to analyze the data.

The test revealed a strong negative correlation between the two variables (Pearson's $r = -0.745$, $p = 1.24e-71$). Given this highly significant p-value, we can confidently conclude that as HDL levels increase, LDL levels tend to decrease. This inverse relationship is an expected result that is supported by existing literature.

```{python, error=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-normality
#| tbl-cap: "Shapiro-Wilk normality test for HDL and LDL"
#| echo: false

## IMPORTING DATA
import pandas as pd
from scipy.stats import shapiro, pearsonr, spearmanr
from pyprojroot import here

df = pd.read_csv(here("data/Part_1/CLEAN_DATA_SET_REFERENCE_EVAL.csv"))
data = df.copy()

## ANALYSIS
HDL = data['HDL'].dropna()
LDL = data['LDL'].dropna()

shapiro_HDL = shapiro(HDL)
shapiro_LDL = shapiro(LDL)

normality_table = pd.DataFrame({
    'Variable': ['HDL', 'LDL'],
    'Statistic': [shapiro_HDL.statistic, shapiro_LDL.statistic],
    'p-value': [shapiro_HDL.pvalue, shapiro_LDL.pvalue],
    'Conclusion': [
        'Reject H0 (non-normal)' if shapiro_HDL.pvalue < 0.05 else 'Fail to reject H0 (normal)',
        'Reject H0 (non-normal)' if shapiro_LDL.pvalue < 0.05 else 'Fail to reject H0 (normal)'
    ]
})

normality_table.style.format({'Statistic': '{:.3f}', 'p-value': '{:.2e}'}).hide(axis='index')
```

```{python}
#| label: tbl-correlation
#| tbl-cap: "Correlation between HDL and LDL"
#| include: false

if shapiro_HDL.pvalue > 0.05 and shapiro_LDL.pvalue > 0.05:
    method = 'Pearson'
    corr, p_value = pearsonr(HDL, LDL)
else:
    method = 'Spearman'
    corr, p_value = spearmanr(HDL, LDL)

correlation_table = pd.DataFrame({
    'Method': [method],
    'Correlation coefficient': [corr],
    'p-value': [p_value]
})

correlation_table.style.format({'Correlation coefficient': '{:.3f}', 'p-value': '{:.2e}'}).hide(axis='index')

```

## 1 Analysis of 1 Continuous Variable vs 1 Categorical Variable.

As discussed before, a initial visual inspection of @fig-ldl-size and @fig-hdl-size suggested an apparent difference in HDL and LDL levels between tall and small individuals, regardless of planet. To statistically corroborate these observations, the data was grouped first by Size, and subsequently by Planet.

To determine the appropriate statistical comparison, each subset was evaluated for normality using the Shapiro-Wilk test. All groups yielded a $p-value \> 0.05$, thereby failing to reject the null hypothesis: data follows a normal distribution. Consequently, parametric independent Welch T-tests were chosen over the non-parametric Kruskal-Wallis test for group comparisons, as detailed in @tbl-normality-groups.

As suspected, the T-test results comparing the mean differences of HDL and LDL grouped by Size (tall vs. small) yielded highly significant p-values ($p \< 0.001$), firmly rejecting the null hypothesis of equal means. Conversely, this significant difference is entirely absent when grouping HDL or LDL by Planet ($p \> 0.05$). Therefore, we can conclude that physical size, rather than planetary origin, is the primary driver of the differences observed among these groups.

```{python, error=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-normality-groups
#| tbl-cap: "Normality of groups"
#| include: false
## IMPORTING DATA
import pandas as pd
from scipy.stats import ttest_ind, shapiro, kruskal

df = pd.read_csv('../data/Part_1/CLEAN_DATA_SET_REFERENCE_EVAL.csv')
data = df.copy()

## Split the data into groups
tall_HDL = data[data['Size'] == 'Tall']['HDL'].dropna()
small_HDL = data[data['Size'] == 'Small']['HDL'].dropna()
tall_LDL = data[data['Size'] == 'Tall']['LDL'].dropna()
small_LDL = data[data['Size'] == 'Small']['LDL'].dropna()
earth_HDL = data[data['Planet'] == 'Earth']['HDL'].dropna()
venus_HDL = data[data['Planet'] == 'Venus']['HDL'].dropna()
earth_LDL = data[data['Planet'] == 'Earth']['LDL'].dropna()
venus_LDL = data[data['Planet'] == 'Venus']['LDL'].dropna()

## Function to check normality
def analyze_normality(variable, variable_name):
    shapiro_test = shapiro(variable)
    if shapiro_test.pvalue < 0.05:
        conclusion = "Reject Ho (not normal)"
    else:
        conclusion = "Fail to reject Ho (normal)"
    return variable_name, shapiro_test.pvalue, conclusion

## Function to compare two groups
def test_analysis(var1, var2, p_var1, p_var2, comparison_name):
    if p_var1 > 0.05 and p_var2 > 0.05:
        # Both distributions are normal, use Welch's t-test
        stat, p_value = ttest_ind(var1, var2, equal_var=False)
        test_used = "T-test (Welch)"
    else:
        # At least one distribution is not normal, use Kruskal-Wallis
        result = kruskal(var1, var2)
        stat, p_value = result.statistic, result.pvalue
        test_used = "Kruskal-Wallis"
    conclusion = "Significant" if p_value < 0.05 else "Not significant"
    return comparison_name, test_used, stat, p_value, conclusion

## Run normality checks
normality_results = [
    analyze_normality(tall_HDL, "Tall-HDL"),
    analyze_normality(small_HDL, "Small-HDL"),
    analyze_normality(tall_LDL, "Tall-LDL"),
    analyze_normality(small_LDL, "Small-LDL"),
    analyze_normality(earth_HDL, "Earth-HDL"),
    analyze_normality(venus_HDL, "Venus-HDL"),
    analyze_normality(earth_LDL, "Earth-LDL"),
    analyze_normality(venus_LDL, "Venus-LDL"),
]

normality_table = pd.DataFrame(normality_results, columns=["Variable", "Shapiro p-value", "Conclusion"])
normality_table.style.format({'Shapiro p-value': '{:.3f}'}).hide(axis='index')
```

```{python}
#| label: tbl-comparison-groups
#| tbl-cap: "Comparison groups"
#| echo: false

## Run comparisons (uses p-values already computed above)
p_tall_HDL = normality_results[0][1]
p_small_HDL = normality_results[1][1]
p_tall_LDL = normality_results[2][1]
p_small_LDL = normality_results[3][1]
p_earth_HDL = normality_results[4][1]
p_venus_HDL = normality_results[5][1]
p_earth_LDL = normality_results[6][1]
p_venus_LDL = normality_results[7][1]

comparison_results = [
    test_analysis(tall_HDL, small_HDL, p_tall_HDL, p_small_HDL, "Tall vs Small (HDL)"),
    test_analysis(tall_LDL, small_LDL, p_tall_LDL, p_small_LDL, "Tall vs Small (LDL)"),
    test_analysis(earth_HDL, venus_HDL, p_earth_HDL, p_venus_HDL, "Earth vs Venus (HDL)"),
    test_analysis(earth_LDL, venus_LDL, p_earth_LDL, p_venus_LDL, "Earth vs Venus (LDL)"),
]

comparison_table = pd.DataFrame(
    comparison_results,
    columns=["Comparison", "Test", "Statistic", "p-value", "Conclusion"]
)

# Shorten long text so the table fits in one column
comparison_table["Test"] = comparison_table["Test"].replace({
    "T-test (Welch)": "T-test",
    "Kruskal-Wallis": "K-W"
})
comparison_table["Conclusion"] = comparison_table["Conclusion"].replace({
    "Significant": "Sig.",
    "Not significant": "N.S."
})

(
    comparison_table.style
    .format({'Statistic': '{:.3f}', 'p-value': '{:.2e}'})
    .hide(axis='index')
    .set_table_styles([
        {'selector': 'th, td', 'props': [('font-size', '11px'), ('padding', '4px 8px')]}
    ])
)
```

# References